### 2020-02-10 submission/abstract_file.tex
full pass

In many practical applications, one works with deep neural networks (DNNs) models trained by someone else.
For such \emph{pretrained models}, one typically does not have access to training data or test data, 
and, moreover,  does not know many detals aout the  model, such as
the specifics of the training data, the loss function, the hyperparameter values, etc.
Given just one or more pretrained models, can one say anything about the expected performance or quality of the models ?
Here, we present and evaluate empirical quality metrics for pretrained neural network models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (the Frobenius and  Spectral norms). as well as
new Power-Law (PL) based metrics, the PL exponent $\alpha$ and the Weighted Alpha ($\hat{\alpha}$) metrics,
from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$ (Shatten) Norm metric.
We find that norm-based metrics correlate well with reported test accuracies of well-trained models
across nearly all CV architecture series.
PL-based metrics can also be used to characterize fine-scale properties of these models, and
we define the layer-wise \emph{Correlation Flow} as new quality assesment.
We also find that PL-based metrics do much better--quantitatively-- at discriminating among
a series of ``good-better-best'' models, and qualitatively better at discriminating ``good-versus-bad'' models.
On the other hand, we find that norm-based metrics can not distinguish ``good-versus-bad'' models--which, 
arguably is the point of needing quality metrics.  Indeed, they may give spurious results.
We show how poorly-trained (and/or poorly-finetuned) models may exhbit both \emph{Scale Collapse}
and unsually large PL exponents $\alpha\gg 6$.  In particular, we show that the \emph{WeightWatcher} tool
can be used to identify when a pretrained DNN has problems that can not be detected simply
by examining the test accuracy.

>>>>

In many practical applications, one works with deep neural network (DNN) models trained by someone else.
For such \emph{pretrained models}, one typically does not have access to training data or test data.
Moreover, one does not know many details about the model, such as the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, can one say anything about the expected performance or quality of the models?
Here, we present and evaluate empirical quality metrics for pretrained DNN models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (Frobenius and Spectral norms) as well as newer Power-Law (PL) based metrics (including fitted PL exponents, $\alpha$, and the Weighted Alpha metric, $\hat{\alpha}$), from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$-Shatten Norm metric.
We find that norm-based metrics correlate well with reported test accuracies for well-trained models across nearly all CV architecture series.
On the other hand, we find that norm-based metrics can not distinguish ``good-versus-bad'' models---which, arguably is the point of needing quality metrics.  
Indeed, they may give spurious results.
We also find that PL-based metrics do much better---quantitatively better at discriminating among a series of ``good-better-best'' models, and qualitatively better at discriminating ``good-versus-bad'' models.
PL-based metrics can also be used to characterize fine-scale properties of these models, and we introduce the layer-wise \emph{Correlation Flow} as new quality assesment.
We show how poorly-trained (and/or poorly fine-tuned) models may exhbit both \emph{Scale Collapse} and unsually large PL exponents, $\alpha\gg 6$, in particular for recent NLP models.
Our techniques, as implemented in the \emph{WeightWatcher} tool, can be used to identify when a pretrained DNN has problems that can not be detected simply by examining training/test accuracies.

### 2020-02-10 submission/all_models.tex
clean up for submission
full pass

\section{Comparison of all pretrained CV Models}
\label{sxn:all_cv_models}

\michael{Now we go back to the CV models and look at over 100 of them, and draw'
more broad conclusions about why alpha and the new norm is better, by looking at the MSE}

Here, we use the Weightwatcher tool to analyze over 100 pretrained computer vision (CV) models Pytorch, including image classification and segmentation models.  They have been pretrained on nine datasets, ImageNet-1K, and CIFAR-10, CIFAR-100, Street View House Numbers (SVHN), Caltech-UCSD Birds-200-2011 (CUB-200-2011), Pascal VOC2012, ADE20K, Cityscapes, and Common Objects in Context (COCO). The pretrained models and their accuracy metrics are summarized in the osmr github \charles{add link}

We provide more details in Appendix~\ref{sxn:appendix}.

In this paper, we propose that the Weightwatcher tool could be used to predict the trends in the generalization accuracy of deep neutral network without a test set. To test our proposition, we choose simple linear regression to analyze the relationship between the Weightwatcher metrics and the traditional accuracy metric obtained with a test set.
% (we avoid polynomial regressions as they are more prone to overfitting and does not make economic sense).
We regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).  These include the Top5 errors for the ImageNet-1K mode, the percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and the Pixel accuracy (Pix.Acc.) and Intersection-Over-Union (mIOU) for XXX.
We regress them individually on each of the Weightwatcher log Norm metrics, as described above.

\begin{table}[t]
\small
\begin{center}
%\begin{tabular}{|p{1in}|c|c|c|c|}
\begin{tabular}{|p{0.75in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &0.55 &0.64 &0.64 \\
$R^{2}$ (std)  & 0.34 &0.36 &0.29 &0.30 \\
\hline
$MSE$ (mean)   & 4.54 &9.62 &3.14 &2.92 \\
$MSE$ (std)    & 8.69 &23.06 &5.14 &5.00 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average log norm metrics across 5 computer vision datasets, 17 Architectures, covering 168 (out of 309) different pretrained DNNs.  We only conclude regressions for architecturs with 4 or more data points, and which are postively correlated with the test error.  These results can be readily reproduced using the Google Colab notebooks (see Appendix)}
\label{table:results}
\end{table}

>>>>

\section{Comparing Hundreds of CV Models}
\label{sxn:all_cv_models}

In this section, we summarize results from a large-scale analysis of hundreds of CV models, including models developed for image classification, segmentation, and a range of related tasks.  
Our aim is to complement the detailed results from Sections~\ref{sxn:cv} and~\ref{sxn:nlp} by providing broader conclusions. % about norm-based and PL-based metrics.
The models we consider have been pretrained on nine datasets, including ImageNet-1K, and CIFAR-10, CIFAR-100, Street View House Numbers (SVHN), Caltech-UCSD Birds-200-2011 (CUB-200-2011), Pascal VOC2012, ADE20K, Cityscapes, and Common Objects in Context (COCO). 
%%The pretrained models and their accuracy metrics are summarized in the osmr github~\cite{charles-add-link}.
We provide more details about our experimental setup in Appendix~\ref{sxn:appendix}.

We choose simple linear regression to analyze the relationship between quality metrics (computed with the \emph{WeightWatcher} tool) and traditional accuracy metrics (publicly-reported from a test set).
% (we avoid polynomial regressions as they are more prone to overfitting and does not make economic sense).
We regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).  
These include Top5 errors for the ImageNet-1K model, percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and Pixel accuracy (Pix.Acc.) and Intersection-Over-Union (IOU) for other models.
We regress them individually on each of the norm-based and PL-based metrics, as described above.

Our results are summarized in Table~\ref{table:results}.
For the mean, larger $R^{2}$ and smaller $MSE$ are desirable; and for the standard deviation, smaller values are desirable.
Taken as a whole, over the entire corpus of data, PL-based metrics are somewhat better for both the $R^{2}$ mean and standard deviation; and PL-based metrics are much better for $MSE$ mean and standard deviation.
These (and other) results suggest our conclusions from Sections~\ref{sxn:cv} and~\ref{sxn:nlp} hold much more generally, and they suggest ovbious questions for future work.

\begin{table}[t]
\small
\begin{center}
%\begin{tabular}{|p{1in}|c|c|c|c|}
\begin{tabular}{|p{0.75in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &  0.55 & \textbf{0.64} & \textbf{0.64} \\
$R^{2}$ (std)  & 0.34 &  0.36 & \textbf{0.29} &          0.30 \\
\hline
$MSE$ (mean)   & 4.54 &  9.62 &          3.14 & \textbf{2.92} \\
$MSE$ (std)    & 8.69 & 23.06 &          5.14 & \textbf{5.00} \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average Log Norm and Weighted Alpha metrics across 5 CV datasets, 17 architectures, covering 168 (out of 309) different pretrained DNNs.  
         We include regressions only for architectures with 4 or more data points, and which are postively correlated with test error.
         These results can be readily reproduced using the Google Colab notebooks (see Appendix~\ref{sxn:appendix}). 
        }
\label{table:results}
\end{table}

### 2020-02-10 submission/background.tex
full pass

Most theory for DNNs is applied to small toy models and assumes access to data.
There is very little work asking how to predict, in a theoretically-principled manner, the quality of large-scale state-of-the-art DNNs, and how to do so without access to training data or testing data or details of the training protocol, etc.
Our work is, however, loosely related to several other lines of work.

@@

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon results from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statisical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

There is also a large body of work on using norm-based metrics to bound generalization error~\cite{NTS15, BFT17_TR, LMBx18_TR}.
In this area, theoretical work aims to prove generalization bounds, and applied work uses these norms as a regularization function to improve training.
While we do find that norms provide relatively good quality metrics, at least for distinguishing ``good-better-best'' among well-trained models, we are not interested in proving generalization bounds or developing new regularizers.

\paragraph{Practical problems need a practical theory.}
\nred{Exiting ML theory does not address many practical AI problems that our techiques directly target.}
%There are many very practical problems in ML that are poorly addressed by existing theory and that either motivated our work or should be addressable by our techniques.
Here are several.
\begin{itemize}[leftmargin=*]
\item
\textbf{Simplicity, or lack of,  accuracy metrics.}
\michael{XXX.  Put in two or at most three sentences.}
\item
\textbf{Information leakage in the production pipeline.}
\michael{XXX.  Put in two or at most three sentences.}
\item
\textbf{Cost of acquring labeled data.}
\michael{XXX.  Put in two or at most three sentences.}
\end{itemize}

\charles{see comments in tex file}\michael{Let's touch base, not sure what to do.}
% we adress this above
%Importantly, there are many examples in ML where (as a practical matter) there is no reliable notion of accuracy, e.g., when generating fake text or realiztic chatbots, developing self driving cars, or just clusterting user profiles.
%\footnote{For example, current chatbots use perplexity as a proxy for passing a Turing test.}
%, and when distilling a reliable model in some way to obtain comparable training/test quality but that damages the model in some other subtle way. 
%\michael{That last example is awkward.  It would be good to have a better example and plant seeds for model distillation elsewhere.}

>>>>

Most theory for DNNs is applied to small toy models and assumes access to data.
There is very little work asking how to predict, in a theoretically-principled manner, the quality of large-scale state-of-the-art DNNs, and how to do so without access to training data or testing data or details of the training protocol, etc.
Our 
approach  %%  work 
is, however, 
%% loosely 
related to 
two  %%  several 
other lines of work.

@@

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statisical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

There is also a large body of work on using norm-based metrics to bound generalization error~\cite{NTS15, BFT17_TR, LMBx18_TR}.
In this area, theoretical work aims to prove generalization bounds, and applied work uses these norms to construct regularizers to improve training.
While we do find that norms provide relatively good quality metrics, at least for distinguishing good-better-best among well-trained models, we are not interested in proving generalization bounds or developing new regularizers.

%% \paragraph{Practical problems need a practical theory.}
%% \nred{Exiting ML theory does not address many practical AI problems that our techiques directly target.}
%% %There are many very practical problems in ML that are poorly addressed by existing theory and that either motivated our work or should be addressable by our techniques.
%% Here are several.
%% \begin{itemize}[leftmargin=*]
%% \item
%% \textbf{Simplicity, or lack of,  accuracy metrics.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Information leakage in the production pipeline.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Cost of acquring labeled data.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \end{itemize}
%% 
%% \charles{see comments in tex file}\michael{Let's touch base, not sure what to do.}
%% % we adress this above
%% %Importantly, there are many examples in ML where (as a practical matter) there is no reliable notion of accuracy, e.g., when generating fake text or realiztic chatbots, developing self driving cars, or just clusterting user profiles.
%% %\footnote{For example, current chatbots use perplexity as a proxy for passing a Turing test.}
%% %, and when distilling a reliable model in some way to obtain comparable training/test quality but that damages the model in some other subtle way. 
%% %\michael{That last example is awkward.  It would be good to have a better example and plant seeds for model distillation elsewhere.}

### 2020-02-10 submission/conclusion.tex
full pass

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers), and our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017}, since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data), but again,  our results are the first to demonstrate this.
It is gratifying, also, that PL exponents can be used to provide fine-scale insight such as rationalizing the flow of correlations through a network.

We conclude with a few thoughts on what a \emph{practical theory} of DNNs should look like.
%
We distinguish between what we will call a
\emph{phenomenological theory}
(that describes empirical relationship of phenomena to each other, in a way which is consistent with fundamental theory, but is not directly derived from that theory)
and what can be called a 
\emph{first principles theory} 
(that is applicable to toy models, but that does not scale up to realistic systems if one includes realistic apsects of realistic systems).
%
For most complex highly-engineered systems (aside from complex AI/ML systems), one \emph{uses} phenomenological theory rather than first principles theory. 
(One does not try to solve the Schr\"odinger equation if one is interested in building a bridge or flying an airplane.)
%
Our results, which are based on our \emph{use} of sophisticated statistical mechanics theory to solve an important practical DNN problems, suggests that this approach should be of interest more generally for those interested in developing a practical DNN theory.
\charles{see latex file}
%\charles{it is the oppposite...ab initio theory scales quite well...it is the spherical cow models of physics and ML that do not scale. What we have is a semi-empirical theory.  We use real theory, but require empirical input, at least in the new stat mech work.  What we have introduced is a phenomenology, which to me is different from a semi-empirical or phenomenological theory  }
\michael{Let's touch base to get this right.}

>>>>

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that this approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

\ncyan{
We conclude with a few thoughts on what a \emph{practical theory} of DNNs should look like.
\michael{MM TO DO.}
%
We distinguish between what we will call a
\emph{phenomenological theory}
(that describes empirical relationship of phenomena to each other, in a way which is consistent with fundamental theory, but is not directly derived from that theory)
and what can be called a 
\emph{first principles theory} 
(that is applicable to toy models, but that does not scale up to realistic systems if one includes realistic apsects of realistic systems).
%
For most complex highly-engineered systems (aside from complex AI/ML systems), one \emph{uses} phenomenological theory rather than first principles theory. 
(One does not try to solve the Schr\"odinger equation if one is interested in building a bridge or flying an airplane.)
%
Our results, which are based on our \emph{use} of sophisticated statistical mechanics theory to solve an important practical DNN problems, suggests that this approach should be of interest more generally for those interested in developing a practical DNN theory.
%\charles{it is the oppposite...ab initio theory scales quite well...it is the spherical cow models of physics and ML that do not scale. What we have is a semi-empirical theory.  We use real theory, but require empirical input, at least in the new stat mech work.  What we have introduced is a phenomenology, which to me is different from a semi-empirical or phenomenological theory  }
}
